Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-08 E-Hypergraphs-LICS24.tex
small updates to S5

CHRIS: REMARK THAT WE CANNOT DEAL WITH SIGNATURES WITH $f:0 \to 0$ MORPHISMS, AND WHY!
As extended cospans have a more general notion of interface we need to make some adjustments to DPOI rewriting.
The first thing is that we need to account for internal interfaces after all.
The main issue is then that internal interfaces may not be preserved during rewriting.
To illustrate this, consider the following rewrite rule
\[\langle x \xrightarrow{} x' \xrightarrow{} \mathcal{L} \xleftarrow{} y' \xleftarrow{} y, x \xrightarrow{} x'' \xrightarrow{} \mathcal{R} \xleftarrow{} y'' \xleftarrow{} y \rangle \]
such that $x' \not \cong x''$ and $y' \not \cong y''$ and consider the following e-hypergraph to perform the rewrite on
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{G} \xleftarrow{} m' \xleftarrow{} m
\] such that there is a matching $m : \mathcal{L} \to \mathcal{G}$.
When removing the occurence of $\mathcal{L}$ from $\mathcal{G}$, e.g. by computing the pushout complement, the internal interface of $\mathcal{G}$ should be modified as the verticies of $\mathcal{L}$ corresponding to strict internal interface of $\mathcal{L}$ (and hence to strict internal interface of $\mathcal{G}$) should be removed from $\mathcal{G}$ and therefore from $\mathcal{G}$'s internal interface.
Then, when gluing $\mathcal{R}$ into the created context, we also introduce strict internal interfaces of $\mathcal{R}$ into this context.
% The input interface on the left-hand side of the rule is $\{1, 2\}$, while the input interface on the right-hand side of the rule is $\{1,2,5,6,7,8\}$.
% Note that the outermost interfaces are preserved.

@@

To account for such changes in the interfaces we put some additional structure on DPOI rewriting for the case of $\catname{EHyp({\Sigma})}$.
Consider a DPOI diagram with extra constraints below which encodes
the notion that e-hypergraph $G$ rewrites into $H$ with external interfaces $n_{ext}+m_{ext}$.
Note the addition of internal interfaces and how we use the same trick to go from cospan $n_{ext} \xrightarrow{} g_i \xrightarrow{} G \xleftarrow{} g_o \xleftarrow{} m_{ext}$ to $G \xleftarrow{} g_{i} + g_{o} \xleftarrow{} n_{ext} + m_{ext}$.
The original DPO square is depicted in blue.
% \[\begin{tikzcd}
% 	&&&& {i + j} \\
% 	{l_i+l_o} && L &&&& R && {r_i+r_o} \\
% 	\\
% 	{g_i+g_o} && {G^{\urcorner}} && C && {^{\ulcorner}H} && {h_i+h_o} \\
% 	\\
% 	&&&& {n+m}
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-7]
% 	\arrow["m", color={rgb,255:red,214;green,92;blue,92}, from=2-3, to=4-3]
% 	\arrow["{[i_c,j_c]}", color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=2-7, to=4-7]
% 	\arrow["{[n_c,m_c]}"', color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-7]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-7]
% 	\arrow[from=2-1, to=2-3]
% 	\arrow["{[f_i,f_o]}", from=2-1, to=4-1]
% 	\arrow[from=4-1, to=4-3]
% 	\arrow[from=2-9, to=2-7]
% 	\arrow[from=4-9, to=4-7]
% 	\arrow["{[h_i,h_o]}"{description}, from=2-9, to=4-9]
% 	\arrow[from=6-5, to=4-9]
% 	\arrow[from=6-5, to=4-1]
% 	\arrow["{[i_l,j_l]}"{description}, from=1-5, to=2-1]
% 	\arrow["{[i_r,j_r]}"{description}, from=1-5, to=2-9]
% \end{tikzcd}\]

@@

For monogamous extended cospans $x_{ext} \xrightarrow{} l_{i} \xrightarrow{} L \xleftarrow{} l_{o} \xleftarrow{} y_{ext}$ and $n_{ext} \xrightarrow{} g_{i} \xrightarrow{} G \xleftarrow{} g_{o} \xleftarrow{} m_{ext}$ and mono $f : L \to G$, a pushout complement $x_{ext} + y_{ext} \to C \to G^{\urcorner}$ as depicted below
\[
\scalebox{0.75}{

>>>>

CHRIS: REMARK THAT WE CANNOT DEAL WITH SIGNATURES WITH $f:0 \to 0$ MORPHISMS, AND WHY!
Because extended cospans have a more general notion of interface,  we need to make some adjustments to DPOI rewriting as presented in Section \ref{sec:combinatorial-semantics}.
The first thing is that we need to account for is the existence of \textit{internal} interfaces,  and in particular that internal interfaces are not necessarily preserved during rewriting.
To illustrate this, consider the following rewrite rule
\[\langle x \xrightarrow{} x' \xrightarrow{} \mathcal{L} \xleftarrow{} y' \xleftarrow{} y, x \xrightarrow{} x'' \xrightarrow{} \mathcal{R} \xleftarrow{} y'' \xleftarrow{} y \rangle \]
such that $x' \not \cong x''$ and $y' \not \cong y''$ and consider the following e-hypergraph to perform the rewrite on
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{G} \xleftarrow{} m' \xleftarrow{} m
\] such that there is a matching $m : \mathcal{L} \to \mathcal{G}$.
When removing the occurence of $\mathcal{L}$ from $\mathcal{G}$, e.g. by computing the pushout complement, the internal interface of $\mathcal{G}$ should be modified.  This is because the verticies of $\mathcal{L}$ corresponding to strict internal interface of $\mathcal{L}$ (and hence to strict internal interface of $\mathcal{G}$) should be removed from $\mathcal{G}$ and therefore from $\mathcal{G}$'s internal interface.
Then, when gluing $\mathcal{R}$ into the created context, we also introduce strict internal interfaces of $\mathcal{R}$ into this context.
% The input interface on the left-hand side of the rule is $\{1, 2\}$, while the input interface on the right-hand side of the rule is $\{1,2,5,6,7,8\}$.
% Note that the outermost interfaces are preserved.

@@

To account for such changes in the interfaces we require additional structure on DPOI rewriting for $\catname{EHyp({\Sigma})}$.
Consider a DPOI diagram with extra constraints below which encodes
the notion that e-hypergraph $G$ rewrites into $H$ with external interfaces $n_{ext}+m_{ext}$.
Note the addition of internal interfaces and how we use the same trick to go from cospan $n_{ext} \xrightarrow{} g_i \xrightarrow{} G \xleftarrow{} g_o \xleftarrow{} m_{ext}$ to $G \xleftarrow{} g_{i} + g_{o} \xleftarrow{} n_{ext} + m_{ext}$.
The original DPO square is depicted in blue.
% \[\begin{tikzcd}
% 	&&&& {i + j} \\
% 	{l_i+l_o} && L &&&& R && {r_i+r_o} \\
% 	\\
% 	{g_i+g_o} && {G^{\urcorner}} && C && {^{\ulcorner}H} && {h_i+h_o} \\
% 	\\
% 	&&&& {n+m}
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-7]
% 	\arrow["m", color={rgb,255:red,214;green,92;blue,92}, from=2-3, to=4-3]
% 	\arrow["{[i_c,j_c]}", color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=2-7, to=4-7]
% 	\arrow["{[n_c,m_c]}"', color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-7]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-7]
% 	\arrow[from=2-1, to=2-3]
% 	\arrow["{[f_i,f_o]}", from=2-1, to=4-1]
% 	\arrow[from=4-1, to=4-3]
% 	\arrow[from=2-9, to=2-7]
% 	\arrow[from=4-9, to=4-7]
% 	\arrow["{[h_i,h_o]}"{description}, from=2-9, to=4-9]
% 	\arrow[from=6-5, to=4-9]
% 	\arrow[from=6-5, to=4-1]
% 	\arrow["{[i_l,j_l]}"{description}, from=1-5, to=2-1]
% 	\arrow["{[i_r,j_r]}"{description}, from=1-5, to=2-9]
% \end{tikzcd}\]

@@

For monogamous extended cospans $x_{ext} \xrightarrow{} l_{i} \xrightarrow{} L \xleftarrow{} l_{o} \xleftarrow{} y_{ext}$ and $n_{ext} \xrightarrow{} g_{i} \xrightarrow{} G \xleftarrow{} g_{o} \xleftarrow{} m_{ext}$ and mono $f : L \to G$, a pushout complement $x_{ext} + y_{ext} \to C \to G^{\urcorner}$ as depicted in the square below
\[
\scalebox{0.75}{
